### 2012-02-21 docs/notes.tex
Implement likelihood that can work on subset of events.  Add test.
Debuggin slice sampler drop offs.
Update draft with some tikz

\pagestyle{empty}
\documentclass[11pt]{article}
\usepackage{graphicx}
\usepackage{verbatim}
\usepackage{url}
\usepackage{graphicx, color}
\usepackage{amsmath}
%\usepackage[small,compact]{titlesec} 
%\usepackage[small,it]{caption}

@@

\abstract{TODO}

@@

Borrowing from the intuition of stochastic blockmodels, we propose a hierarchical model of continuous time network data that learns latent clusters of nodes which share similar patterns of interaction.  [TODO: Explain how our experiments show that this helps.]   [TODO: perhaps use ``role'' terminology.]

@@

\begin{figure}
 \def\svgwidth{6in}
  \input{example1.pdf_tex}
\caption{Illustration of event data and the assumptions of the model.  Left: An sequence of two events among four nodes: (1,2) occurs at time $t_1$ and (2,3) occurs at time $t_2$.  Center: The intensity functions $\lambda_{1,2}(t)$ (solid) and $\lambda_{34}(t)$ (dotted).  Note $\lambda_{34}$ can only change when events occur involving either node 3 or node 4.  Right: Illustration of the block model, where events in cluster 1 (circles) are governed by the vector $\beta_{1,1}$, events in cluster 2 (squares) are governed by $\beta_{2,2}$, and an circle-to-square event governed by $\beta_{1,2}$.}
\label{fig:example}
%\includegraphics[width=6in]{example}
\end{figure}

@@

\begin{itemize}
%\item Classroom:  Perhaps within a couple of the classrooms we can infer subgroups who tend to obey different relational event dynamics.
\item Email: In small organizations there may be groups of people who tend to have a similar style of interaction, and who also share a pattern of interaction with those outside the group or in another group.
%\item Citation networks: Perhaps some groups of papers obey some types of rich-get-richer effects more than others.
\item We collected tweets from Twitter.com occurring between from May 11, 2009 to January 26, 2012 that contained the hashtag \texttt{\#rstats}.  This hashtag is used to denote messages pertaining to the R statistical computing environment and sometimes statistical discussion more generally.  We collect dyadic events by selecting tweets beginning with the \texttt{@} symbol (called a \emph{mention}), and mark the first mentioned user as the recipient.  Of 28337 total tweets in this time period, 3926 were directed events among a total of 1079 users.  We use the subset of 487 users who participated in more than one event, using a training set of 2000 events and a test set of 1330 events.
\end{itemize}

@@

Several baselines are included for comparison: \texttt{uniform} places uniform probability on all possible dyads, \texttt{online} ranks events at time $t$ by the number of times the dyad has occurred previously $\sum_{m:t_m < t} I(i_m=i,j_m=j)$, and \texttt{marginal} uses the product of the observed marginal frequencies $\sum_{m:t_m < t} I(i_m=i) \sum_{m:t_m < t} I(j_m=j)$.  Note for processes that are homogeneous over time, \texttt{online} should do well with large amounts of data while \texttt{marginal} should roughly model heterogeneity in activity among individuals.

@@

@@

\subsection{Extensions}
Possible future work could extend the present framework to a mixed-membership model where for each effect there is a latent cluster assignment.  This way a group of nodes might contact high in-degree nodes at a similar rate, though they vary in the degree to which they adhere to reciprocity.  As in the mixed-membership stochastic blockmodel \cite{Airoldi2008,Shafiei2010}, one could assume that each node has vector of probabilities, $\boldsymbol{\pi}_i$, for being assigned to each of the latent clusters.  In that case $z_i \sim \mbox{Categorical}(\boldsymbol{\pi}_i)$ for each node $i$ and $j$, and the ties are modeled via a stochastic blockmodel $\mbox{logit} p(y_{ij}) = \eta_{z_i,z_j}$.  One could extend our current framework by assuming that for each event $m$, any node  $i,j \in \mathcal{R}_{i_m,j_m}$ redraws their latent class.

@@

>>>>

\pagestyle{empty}
\documentclass[11pt]{article}
\usepackage{tikz}
\tikzstyle{vertex}=[auto=left,circle,fill=black!25,minimum size=20pt,inner sep=0pt]

%\usepackage{tikz,fullpage}
%\usetikzlibrary{arrows,petri,topaths}%
%\usepackage{tkz-berge}
\usepackage{verbatim}
\usepackage{url}
\usepackage{graphicx, color}
\usepackage{amsmath}
\usepackage{subfigure}
%\usepackage[small,compact]{titlesec} 
%\usepackage[small,it]{caption}

@@

\abstract{

\begin{itemize}
\item Blockmodels and clustering algorithms often applied to static networks for exploratory data analysis
\item continuous time methods allow one to predict future events given the previous history
\item we combine these two ideas with a hierarchical model such that each pair of blocks has a set of parameters guiding the event dynamics for those dyads
\item prediction experiments on real datasets
\item show that the parameter estimates are interpretable
\end{itemize}

}

@@

Borrowing from the intuition of stochastic blockmodels, we propose a hierarchical model of continuous time network data that learns latent clusters of nodes which share similar patterns of interaction.  The behavior of the model is illustrated with data simulated from the model.  Using  data sets of dyadic communication among people, we demonstrate the ability to improve upon simple baselines for predicting the next event.  Finally we show that the parameter estimates provide an interpretable structure to the event dynamics among blocks of individuals, sometimes identifying particular \emph{roles} of communication that exist within the network.

@@

\begin{figure}

\centering 
\subfigure[Dynamic network data]
{
\begin{tikzpicture}[scale=.8,transform shape,thick]
  \node[vertex] (n1) at (0,0)  {1};
  \node[vertex] (n2) at (2,2)  {2};
  \node[vertex] (n3) at (4,0) {3};
  \node[vertex] (n4) at (2,-2) {4};
\draw[-stealth] (.4,.4) -- (1.6,1.6);
\draw[-stealth] (2.4,1.6) -- (3.6,.4);
\node at (.7,1.1) {$t_1$};
\node at (3.3,1.1) {$t_2$};
\end{tikzpicture}
}
\subfigure[Intensities for two dyads]
{
\begin{tikzpicture}[scale=1,transform shape,thick]
\draw[-stealth] (0,0) -- (0,3);
\draw[-stealth] (0,0) -- (5,0);

% First timeline
\draw (0,2) -- (1,2);
\draw[fill=black] (1,2) circle (0.75mm);
\draw (1,1.7) circle (0.75mm);
\draw (1.08,1.7) -- (3,1.7);
\draw[fill=black] (3,1.7) circle (0.75mm);
\draw (3,1.7) circle (0.75mm);
\draw[fill=black] (3,2.2) circle (0.75mm);
\draw[-stealth] (3,2.2) -- (5,2.2);

% second timeline
\draw[dashed] (0,1) -- (3,1);
\draw[fill=black] (3,1) circle (0.75mm);
\draw (3,.75) circle (0.75mm);
\draw[dashed, -stealth] (3.08,.75) -- (5,.75);

% labels
\node at (-.5,2) {$\lambda_{1,2}$};
\node at (-.5,1) {$\lambda_{3,4}$};
\node at (1,-.3) {$t_1$};
\node at (3,-.3) {$t_2$};
\end{tikzpicture}
}
\caption{Illustration of event data and the assumptions of the model.  Left: An sequence of two events among four nodes: (1,2) occurs at time $t_1$ and (2,3) occurs at time $t_2$.  Center: The intensity functions $\lambda_{1,2}(t)$ (solid) and $\lambda_{34}(t)$ (dotted).  Note $\lambda_{34}$ can only change when events occur involving either node 3 or node 4. [TODO? Illustration of the block model, where events in cluster 1 (circles) are governed by the vector $\beta_{1,1}$, events in cluster 2 (squares) are governed by $\beta_{2,2}$, and an circle-to-square event governed by $\beta_{1,2}$.]}
\label{fig:example}
\end{figure}

% \begin{figure}
%  \def\svgwidth{6in}
%   \input{example1.pdf_tex}
% \caption{Illustration of event data and the assumptions of the model.  Left: An sequence of two events among four nodes: (1,2) occurs at time $t_1$ and (2,3) occurs at time $t_2$.  Center: The intensity functions $\lambda_{1,2}(t)$ (solid) and $\lambda_{34}(t)$ (dotted).  Note $\lambda_{34}$ can only change when events occur involving either node 3 or node 4.  Right: Illustration of the block model, where events in cluster 1 (circles) are governed by the vector $\beta_{1,1}$, events in cluster 2 (squares) are governed by $\beta_{2,2}$, and an circle-to-square event governed by $\beta_{1,2}$.}
% \label{fig:example}
% %\includegraphics[width=6in]{example}
% \end{figure}

@@

\begin{itemize}
%\item Email: In small organizations there may be groups of people who tend to have a similar style of interaction, and who also share a pattern of interaction with those outside the group or in another group.
\item Tweets from Twitter.com occurring between from May 11, 2009 to January 26, 2012 that contained the hashtag \texttt{\#rstats}.\footnote{Will be made publicly available.}  This hashtag is used to denote messages pertaining to the R statistical computing environment and sometimes statistical discussion more generally.  We collect dyadic events by selecting tweets beginning with the \texttt{@} symbol (called a \emph{mention}), and mark the first mentioned user as the recipient.  Of 28337 total tweets in this time period, 3926 were directed events among a total of 1079 users.  We use the subset of 487 users who participated in more than one event, using a training set of 2000 events and a test set of 1330 events.
\item The Irvine data set
\end{itemize}

@@

Several baselines are included for comparison: \texttt{uniform} places uniform probability on all possible dyads, \texttt{online} ranks events at time $t$ by the number of times the dyad has occurred previously $r_{online}(i,j) = \sum_{m:t_m < t} I(i_m=i,j_m=j)$, and \texttt{marginal} uses the product of the observed marginal frequencies $r_{marg} = \sum_{m:t_m < t} I(i_m=i) \sum_{m:t_m < t} I(j_m=j)$.  Note for processes that are homogeneous over time, \texttt{online} should do well with large amounts of data while \texttt{marginal} should roughly model heterogeneity in activity among individuals.

Our method jointly models both which dyads occur and when they occur.  To compare the above baselines to our model using the likelihood of observed data, we assume each dyad is a Poisson process with estimated  rate $\hat{\lambda}_{i,j} = \frac{M}{t_M} \frac{r_{b}(i,j) + \xi}{\sum_{ij} r_{b}(i,j) + \xi}$, where $r_b(i,j)$ is the statistic a baseline (described above) and $\xi=1$ is a smoothing parameter.

@@

TODO: discuss non-specification of rate and just predicting next
event, could still use our approach?

@@

\subsection{Extensions}
Possible future work could extend the present framework to a mixed-membership model where for each effect there is a latent cluster assignment.  This way a group of nodes might contact high in-degree nodes at a similar rate, though they vary in the degree to which they adhere to reciprocity.  As in the mixed-membership stochastic blockmodel \cite{Airoldi2008,Shafiei2010}, one could assume that each node has vector of probabilities, $\boldsymbol{\pi}_i$, for being assigned to each of the latent clusters.  In that case $z_i \sim \mbox{Categorical}(\boldsymbol{\pi}_i)$ for each node $i$ and $j$, and the ties are modeled via a stochastic blockmodel $\mbox{logit} p(y_{ij}) = \eta_{z_i,z_j}$.  One could extend our current framework by assuming that for each event $m$, any node  $i,j \in \mathcal{R}_{i_m,j_m}$ redraws their latent class.  TODO: Discuss why we *didn't* do this.

@@

\section{Related work}
\cite{Tillman2011} explores the idea of having nonpoisson processes at each edge and solves the master equation for the dynamical system.
